---
title: "Cat Pricing Cluster Blueprint — HTML Report"
author: "Ruchi Malhotra"
output:
  html_document:
    mathjax: null
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5, fig.align = "center")

set.seed(42)

```  


## 1. Libraries
```{r}
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(purrr); library(scales)
  library(ggplot2); library(terra); library(sf); library(cluster)
  library(data.table); library(kableExtra); library(here)
})

options(kableExtra.auto_format = FALSE)

# helper for PML/TVaR (used later)
pml_tvar <- function(x, alpha = 0.99){
  q <- as.numeric(quantile(x, probs = alpha, na.rm = TRUE))
  tail_vals <- x[x >= q]
  tvar <- if (length(tail_vals)) mean(tail_vals) else q
  list(pml = q, tvar = tvar)
}


# helper function for no-wrap column headers
nowrap <- function(x) sprintf("<span style='white-space:nowrap;'>%s</span>", x)

# format $ with commas
fmt_dollar0 <- function(x) scales::comma(x, accuracy = 1)

# color negatives (green = savings), positives red
cell_color_delta <- function(x){
  ifelse(x < 0,
         cell_spec(fmt_dollar0(x), color = "white", background = "#2E7D32", bold = TRUE),
         ifelse(x > 0,
                cell_spec(fmt_dollar0(x), color = "white", background = "#C62828", bold = TRUE),
                fmt_dollar0(x)))
}

```


## 2. Inputs and exposure
```{r, include=FALSE}
hazard_2014_path <- here::here("data","chirps-v3.0.2014.tif")
hazard_2024_path <- here::here("data","chirps-v3.0.2024.tif")
out_dir <- here::here("outputs"); if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
stopifnot(file.exists(hazard_2014_path), file.exists(hazard_2024_path))

n_assets <- 500
exposure <- tibble(
  asset_id = 1:n_assets,
  lat = runif(n_assets, 32.5, 42),
  lon = runif(n_assets, -124.5, -114),
  insured_value = runif(n_assets, 200000, 1000000),
  construction = sample(c("wood","masonry","steel"), n_assets, replace = TRUE),
  occupancy    = sample(c("residential","commercial"), n_assets, replace = TRUE),
  deductible   = 5000,
  limit        = 750000
)

asset_pts <- terra::vect(exposure, geom = c("lon","lat"), crs = "EPSG:4326")
hazard_2014 <- rast(hazard_2014_path)
hazard_2024 <- rast(hazard_2024_path)

exposure$precip_2014 <- terra::extract(hazard_2014, asset_pts)[[2]]
exposure$precip_2024 <- terra::extract(hazard_2024, asset_pts)[[2]]
exposure <- exposure %>%
  mutate(precip_change   = precip_2024 - precip_2014,
         fire_risk_score = log1p(pmax(0, precip_change)))


```

## 3. Clustering (k-means)

```{r}

df_cluster <- exposure %>%
  mutate(construction_num = as.numeric(factor(construction)),
         occupancy_num    = as.numeric(factor(occupancy))) %>%
  select(insured_value, precip_2024, fire_risk_score, construction_num, occupancy_num) %>%
  drop_na()

k <- 5
km_model <- kmeans(df_cluster, centers = k, nstart = 20)
exposure$cluster <- factor(km_model$cluster)

```


## 4. Stochastic events, policy terms, AAL

```{r}
vuln_lookup <- c(wood = 0.6, masonry = 0.4, steel = 0.2)
exposure$vuln_factor <- as.numeric(vuln_lookup[exposure$construction])

n_events <- 1000
event_catalog <- tibble(event_id = 1:n_events,
                        severity = rexp(n_events, rate = 1/1.5))

sim_matrix <- tidyr::expand_grid(asset_id = exposure$asset_id, event_id = event_catalog$event_id) %>%
  left_join(exposure, by = "asset_id") %>%
  left_join(event_catalog, by = "event_id") %>%
  mutate(
    raw_loss   = insured_value * vuln_factor * severity * fire_risk_score * 0.01,
    gross_loss = pmax(0, raw_loss - deductible),
    net_loss   = pmin(gross_loss, limit)
  )

aal_df <- sim_matrix %>%
  group_by(asset_id) %>%
  summarise(cat_load = mean(net_loss, na.rm = TRUE), .groups = "drop")
exposure <- exposure %>% left_join(aal_df, by = "asset_id")

base_rate <- 0.02
construction_factors <- c(wood = 1.2, masonry = 1.0, steel = 0.9)
exposure <- exposure %>%
  mutate(expected_premium = insured_value * base_rate * construction_factors[construction])

```


## 5. Cluster plots and summary
```{r}

cluster_summary <- exposure %>%
  group_by(cluster) %>%
  summarise(
    count                = n(),
    avg_insured_value    = mean(insured_value),
    avg_fire_risk        = mean(fire_risk_score),
    avg_cat_load         = mean(cat_load, na.rm = TRUE),
    total_cat_load       = sum(cat_load, na.rm = TRUE),
    avg_expected_premium = mean(expected_premium),
    avg_loss_ratio       = mean(cat_load / expected_premium, na.rm = TRUE),
    .groups = "drop"
  )

cluster_summary %>%
  mutate(across(where(is.numeric), ~round(., 2))) %>%
  kable(caption = "Cluster-Level Summary of CAT Load and Premium Metrics") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)


```

```{r}
# Make long data for plotting
cluster_summary_long <- cluster_summary %>%
  dplyr::select(cluster, avg_cat_load, avg_expected_premium) %>%
  tidyr::pivot_longer(
    cols = c(avg_cat_load, avg_expected_premium),
    names_to = "Metric", values_to = "Amount"
  ) %>%
  dplyr::mutate(
    Metric = dplyr::recode(Metric,
      avg_cat_load = "CAT Load",
      avg_expected_premium = "Expected Premium"
    )
  )

```

# Plot 1: Average CAT Load vs Expected Premium 
```{r}
ggplot(cluster_summary_long, aes(x = cluster, y = Amount, fill = Metric)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Average CAT Load vs Expected Premium by Cluster",
       x = "Cluster", y = "Amount ($)", fill = "Metric") +
  theme_minimal()

```

# Plot 2: Total CAT Load by Cluster 
```{r}
ggplot(cluster_summary, aes(cluster, total_cat_load, fill = cluster)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Total CAT Load by Cluster", x = "Cluster", y = "Total CAT Load ($)") +
  theme_minimal()

```



# Plot 3: Average Loss Ratio (CAT Load / Premium) by Cluster 
```{r}
ggplot(cluster_summary, aes(cluster, avg_loss_ratio, fill = cluster)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Average Loss Ratio (CAT Load / Premium) by Cluster", x = "Cluster", y = "Loss Ratio") +
  theme_minimal()

```


## 6. EP curves, PML(100)/PML(200), TVaR(99/99.5)

```{r}
dt <- as.data.table(sim_matrix)
event_cluster_losses <- dt[, .(event_loss = sum(net_loss, na.rm=TRUE)),
                           by = .(cluster, event_id)]

metrics <- event_cluster_losses[, {
  m99  <- pml_tvar(event_loss, 0.99)
  m995 <- pml_tvar(event_loss, 0.995)
  .(PML_100 = m99$pml, TVaR_99 = m99$tvar,
    PML_200 = m995$pml, TVaR_99_5 = m995$tvar)
}, by = cluster]

```

```{r}

metrics %>%
  mutate(across(where(is.numeric), ~round(., 0))) %>%
  kable(caption = "PML & TVaR by Cluster") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

```{r}
probs <- seq(0.001, 0.99, by = 0.001)
ep <- event_cluster_losses[, .(
  loss = quantile(event_loss, probs = 1 - probs, na.rm = TRUE),
  ep   = probs
), by = cluster]

ggplot(ep, aes(ep, loss, color = cluster)) +
  geom_line(linewidth = 1.05) +
  scale_x_reverse(labels = scales::percent) +
  labs(title = "EP Curves by Cluster", x = "Exceedance Probability", y = "Loss ($)") +
  theme_minimal()

```

## 7. EP curves after hardening

```{r harden_prep, include = FALSE}

# Option A: global % reduction of vulnerability for all assets
hardening_reduction <- 0.20               # 20% reduction
exposure_hardened <- exposure %>%
  mutate(vuln_factor = vuln_factor * (1 - hardening_reduction))



# ---- Reuse SAME event catalog for comparability ----
# n_events and event_catalog already defined in baseline; we reuse event_catalog

# ---- Re-run stochastic losses with policy terms (hardened) ----
sim_matrix_hardened <- tidyr::expand_grid(asset_id = exposure_hardened$asset_id,
                                          event_id = event_catalog$event_id) %>%
  left_join(exposure_hardened, by = "asset_id") %>%
  left_join(event_catalog,    by = "event_id") %>%
  mutate(
    raw_loss_h   = insured_value * vuln_factor * severity * fire_risk_score * 0.01,
    gross_loss_h = pmax(0, raw_loss_h - deductible),
    net_loss_h   = pmin(gross_loss_h, limit)
  )

# ---- Aggregate event losses at cluster event totals (hardened) ----
event_cluster_losses_h <- as.data.table(sim_matrix_hardened)[,
  .(event_loss = sum(net_loss_h, na.rm = TRUE)),
  by = .(cluster, event_id)
]

# Aggregate to PORTFOLIO-EVENT totals (baseline and hardened)
event_portfolio_losses    <- as.data.table(sim_matrix)[, .(event_loss = sum(net_loss,   na.rm=TRUE)), by = event_id]
event_portfolio_losses_h  <- as.data.table(sim_matrix_hardened)[, .(event_loss = sum(net_loss_h, na.rm=TRUE)), by = event_id]

```

## Portfolio-level EP curve overlay

```{r ep_overlay_portfolio, echo=FALSE, message=FALSE, warning=FALSE}
probs <- seq(0.001, 0.99, by = 0.001)

# Build portfolio EP curves for both scenarios
ep_portfolio <- tibble(
  scenario = "Baseline",
  ep = probs,
  loss = quantile(event_portfolio_losses$event_loss, probs = 1 - probs, na.rm = TRUE)
) %>%
  bind_rows(tibble(
    scenario = "Hardened",
    ep = probs,
    loss = quantile(event_portfolio_losses_h$event_loss, probs = 1 - probs, na.rm = TRUE)
  ))

# Plot overlay
ggplot(ep_portfolio, aes(ep, loss, linetype = scenario)) +
  geom_line(linewidth = 1.05) +
  scale_x_reverse(labels = scales::percent) +
  labs(title = "Impact of Hardening on Tail Losses — Portfolio EP Curves",
       x = "Exceedance Probability", y = "Loss ($)", linetype = "Scenario") +
  theme_minimal()

```

## Portfolio metrics before vs after
```{r portfolio_metrics_compare, echo=FALSE, message=FALSE, warning=FALSE}
# AAL = mean of event losses
aal_base <- mean(event_portfolio_losses$event_loss, na.rm = TRUE)
aal_hard <- mean(event_portfolio_losses_h$event_loss, na.rm = TRUE)

# PML/TVaR helper already defined: pml_tvar()
m99_base  <- pml_tvar(event_portfolio_losses$event_loss, 0.99)
m995_base <- pml_tvar(event_portfolio_losses$event_loss, 0.995)
m99_h     <- pml_tvar(event_portfolio_losses_h$event_loss, 0.99)
m995_h    <- pml_tvar(event_portfolio_losses_h$event_loss, 0.995)

portfolio_compare <- tibble(
  Metric = c("AAL", "PML(100)", "TVaR(99)", "PML(200)", "TVaR(99.5)"),
  Baseline = c(aal_base, m99_base$pml, m99_base$tvar, m995_base$pml, m995_base$tvar),
  Hardened = c(aal_hard, m99_h$pml,   m99_h$tvar,   m995_h$pml,   m995_h$tvar)
) %>%
  mutate(
    `Δ ($)`   = Hardened - Baseline,
    `Δ (%)`   = 100 * (Hardened / Baseline - 1)
  )

portfolio_compare %>%
  mutate(across(c(Baseline, Hardened, `Δ ($)`), ~round(.,0)),
         `Δ (%)` = scales::percent(`Δ (%)`/100, accuracy = 0.1)) %>%
  kable(caption = "Portfolio Metrics — Before vs After Hardening") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

## Cluster-level AAL and tail benefits

```{r cluster_benefits_tbl, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
# Baseline cluster AAL
aal_cluster_base <- as.data.table(sim_matrix)[, .(AAL = mean(net_loss, na.rm=TRUE)), by = cluster]
# Hardened cluster AAL
aal_cluster_h    <- as.data.table(sim_matrix_hardened)[, .(AAL = mean(net_loss_h, na.rm=TRUE)), by = cluster]

# Tail metrics per cluster
metrics_base <- event_cluster_losses[, {
  m99 <- pml_tvar(event_loss, 0.99); m995 <- pml_tvar(event_loss, 0.995)
  .(PML_100 = m99$pml, TVaR_99 = m99$tvar, PML_200 = m995$pml, TVaR_99_5 = m995$tvar)
}, by = cluster]

metrics_h <- event_cluster_losses_h[, {
  m99 <- pml_tvar(event_loss, 0.99); m995 <- pml_tvar(event_loss, 0.995)
  .(PML_100 = m99$pml, TVaR_99 = m99$tvar, PML_200 = m995$pml, TVaR_99_5 = m995$tvar)
}, by = cluster]

cluster_compare <- aal_cluster_base %>%
  rename(AAL_base = AAL) %>%
  full_join(aal_cluster_h %>% rename(AAL_h = AAL), by = "cluster") %>%
  left_join(metrics_base %>% rename_with(~paste0(.x,"_base"), -cluster), by = "cluster") %>%
  left_join(metrics_h %>%    rename_with(~paste0(.x,"_h"),    -cluster), by = "cluster") %>%
  mutate(
    AAL_delta   = AAL_h - AAL_base,
    PML100_dlt  = PML_100_h - PML_100_base,
    TVaR99_dlt  = TVaR_99_h - TVaR_99_base
  ) %>%
  arrange(AAL_delta)

cluster_benefits_tbl <- cluster_compare %>%
  transmute(
    Cluster          = as.character(cluster),
    AAL_Base         = AAL_base,
    AAL_Hardened     = AAL_h,
    dAAL             = AAL_h - AAL_base,
    dPML100          = PML100_dlt,
    dTVaR99          = TVaR99_dlt
  ) %>%
  arrange(dAAL) %>%
  mutate(
    across(c(AAL_Base, AAL_Hardened, dAAL, dPML100, dTVaR99), fmt_dollar0)
  )

cluster_benefits_tbl %>%
  kable(
    format  = "html",
    caption = "Cluster-Level Benefits — AAL and Tail Metrics (Hardened minus Baseline)",
    align   = c("c","r","r","r","r","r"),
    escape  = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped","hover","condensed"),
    full_width = FALSE,
    position   = "left",
    font_size  = 12
  ) %>%
  column_spec(1, width = "6em",  bold = TRUE) %>%
  column_spec(2:6, width = "11em")

```


## AAL savings by cluster bar chart

```{r aal_savings_plot, echo=FALSE, message=FALSE, warning=FALSE}
cluster_compare %>%
  mutate(`AAL Savings ($)` = -AAL_delta) %>%      # negative delta -> savings
  ggplot(aes(x = reorder(cluster, `AAL Savings ($)`), y = `AAL Savings ($)`)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "AAL Savings by Cluster — Effect of Hardening",
       x = "Cluster", y = "Savings ($)") +
  theme_minimal()
```


```{r pricing_adequacy_tbl, results = 'asis', echo=FALSE, message=FALSE, warning=FALSE}

pricing_print <- cluster_summary %>%
  transmute(
    Cluster              = as.character(cluster),
    Avg_CAT_Load         = avg_cat_load,
    Avg_Premium          = avg_expected_premium,
    CAT_Load_to_Premium  = avg_cat_load / avg_expected_premium
  ) %>%
  arrange(desc(CAT_Load_to_Premium)) %>%
  mutate(
    across(c(Avg_CAT_Load, Avg_Premium), ~ fmt_dollar0(.x)),
    CAT_Load_to_Premium = round(CAT_Load_to_Premium, 3)
  )

```


```{r results='asis', echo=FALSE, message=FALSE, warning=FALSE}

pricing_print %>%
  kable(
    format  = "html",
    caption = "Pricing Adequacy by Cluster (CAT Load vs Premium)",
    align   = c("c","r","r","r"),
    escape  = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped","hover","condensed"),
    full_width = FALSE,
    position   = "left",
    font_size  = 12
  ) %>%
  column_spec(1, width = "6em",  bold = TRUE) %>%
  column_spec(2:4, width = "10em")


```

# Interpretation of Results

The results provide a structured view of how clustered exposures perform under simulated catastrophe shocks:


1) **Avg CAT Load vs Expected Premium (Plot 1)**  
   - In every cluster, **Expected Premium > Avg CAT Load**.  
   - That means the modeled catastrophe portion of loss is **covered by the current premium** on an average-year basis.

2) **Total CAT Load by Cluster (Plot 2)**  
   - Cluster **4** contributes the **largest absolute CAT dollars** (portfolio accumulation).  
   - Even with adequate average pricing, high dollar accumulation makes Cluster 4 the **dominant driver of portfolio volatility and tail risk**.

3) **Average Loss Ratio (Plot 3)**  
   - CAT loss ratios (CAT Load / Premium) are **< 1.0** across the board.  
   - Cluster 4, then 2 and 1, sit at the higher end of ratios; Cluster 3 is the lowest.  
   - Interpretation: pricing **isn’t under** for CAT, but some clusters **consume more of the premium** to cover CAT risk, leaving less room for non-CAT perils, expenses, and margin.

4) **PML & TVaR Table**  
   - Cluster **4** shows the **largest PML(100) and TVaR(99)** values → highest **extreme-event exposure**.  
   - TVaR being meaningfully above PML confirms a **fat tail**; this is where reinsurance and capital decisions bite.

5) **EP Curves**  
   - Cluster 4’s curve sits **highest/right-most** in the tail → at the same exceedance probability, it expects **bigger losses** than other clusters.  
   - Clusters 1 and 2 follow; Cluster 3 is most resilient.

---

## Actionable Takeaways

- **Risk transfer & capital:** Size reinsurance layers primarily off **Cluster 4** (tail-heavy), then 2 and 1. Use PML/TVaR to set attachment and limits.  
- **Portfolio steering:** If you want to reduce tail, **de-risk Cluster 4** (blend of underwriting appetite, limits, and accumulation caps).  
- **Resilience capex:** Prioritize hardening for assets in Cluster 4 (and next 2/1). Even if pricing covers average CAT, **tail losses** dominate capital and rating-agency views.  
- **Pricing adequacy (CAT slice only):** Current premium covers modeled CAT across clusters. Any overall pricing action should consider **non-CAT perils + expenses + target margin** on top of the CAT slice.

*Bottom line:* CAT pricing looks **adequate on average**, but **tail concentration** (PML/TVaR, EP curves) makes **Cluster 4** the lever for reinsurance, capital efficiency, and resilience ROI.